evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks

### 2020-07-23 manuscript.tex
Gut

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure1.pdf}
\caption{Explanatory Figures. 
(a) Left: For two neurons (x and y-axis), illustration of sample (dots) and boundary of underlying (circle) first spike stimulus-response distributions for two hypothetical stimuli (red and gold). Right: Illustration of corresponding relative first spike differences. 
(b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
(c) Illustration of boundary of factor-dependent first spike response distribution for a single factor value. The green line shows the mean points that would be predicted for a continuous set of factor values.
(d) Illustration of superposition of two single spike pattern feature representations.
(e) Illustration of `stationary clustered time-warped precise spike timing'. Left: Responses form several clear clusters. The main green cluster is correlated with non-unitary gradient (i.e. different from $45^{\circ}$). Under the assumption of stationarity, the green line and circle are the same as in (c). The orange ellipses illustrate the boundary of an underlying response distribution under the assumption of independence and normality for the sample (solid) and another stimulus (dashed). Centre: Illustration of spike time differences for values of the factor. The green circle illustrates the boundary of a factor-dependent distribution of spike time differences. Right: Conceptual illustration of time warp relative differences.}
\label{fig:fig1}
\end{figure}

@@

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}. The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}. Such first spike patterns allow direct study of the relationship between known stimuli and neuronal activity. A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation of information.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units. That is, the first spike times of two units were plotted against each other for trials on which they both spiked in response to a particular stimulus. The reader is strongly recommended to view these plots in the top left of Supplementary Vid. 1. Simple observation of these plots (Ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs may correlated for pairs of neurons on a subset of trials of a stimulus. Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).

@@

A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing. Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.

A novel algorithm extracted the observed clusters and found that they were correlated appeared above chance levels. To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity. The cluster spike times of both units were stationary for X\% of correlated clusters. For the remaining clusters, which had at least one unit with non-stationary cluster spike times, time series analysis was applied to the cluster spike times of the non-stationary unit/units to remove the non-stationarities (i.e. neural adaptive trend). X\% of the original clusters

The characterisation offers a prediction of the encoding resolution of the possible state-dependent coding form, which would surpass previous unconditioned coding forms and controls for neural adaptive trends. It also provides a precise understanding of stimulus responses and correlative variability, which can be interpreted in the classic sense [Vicente Refs].

The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. Testing the state-dependent decoding performance of the uncovered response properties would require explicit knowledge of the underlying state so is left for future studies.
Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.

@@

The paper uncovers interesting correlative properties, which offer a novel description of neural representations and are suggestive (but not confirmatory) of an internal continuous state varying responses from trial-to-trial.

@@

\textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?

Firstly, for repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity), are the first spike times of pairs of units correlated? This analysis should take into account outliers and the possibility that responses come .

In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels. Clustering was necessary for the detection of a large proportion of correlations.

\textit{2.} For sets of correlated spike pairs, are the angles of correlation significantly different from $45^{\circ}$ (i.e. is the gradient of correlation different from 1)? This question is of interest for two reasons. Firstly such . This leads to the

\textit{3.} Does this offer a potential improvement in encoding capacity

\textit{4.} For pairs of units in which

we addressed the following questions / hypotheses. (1) …. (2) …. (3)…

To our knowledge this is the first report demonstrating that

Cluster spike times of both units were stationary for X\%\todo[color=red!40]{Fill} of correlated clusters. X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying factor co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig. 1d). Estimated factor-conditioned response distributions for these stationary clusters showed a high potential encoding capacity. 
% Algorithms for detecting correlated clusters and estimating correlation angles were developed for this study and verified on surrogate control datasets.

Time series analysis was applied to the non-stationary units of the remaining correlated clusters, to remove neural adaptive trends. A large proportion of detrended clusters were correlated with non-unitary gradient, suggestive of an underlying factor with an improved encoding potential. The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and plasticity. In this study, such consideration gave an estimate of the potential resolution of neural stimulus representation. 
% Although the study analyses responses to single stimuli in isolation, it also considers the proposed and previously suggested neural codes with respect to feature-binding and continua encoding.

The characterised response distribution structure is referred to as \textbf{stationary and non stationary clustered time-warped spike timing}. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons needn't respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a factor-dependent manner.

@@

The dataset used for this paper is reused from an earlier barrel cortex study, of which analyses and data collection methods have previously been reported \cite{reyes2014laminar}. In summary, extracellular recordings were made in 15 anaesthetised Wistar rats using 16 or 128 electrode silicon probes, inserted perpendicularly into the barrel cortex. Single whisker deflections were made at various frequencies (0.1-10Hz) in blocks, each containing typically 200 trials (sweeps) with an intertrial interval of 5 s (occasionally 10, 15, or 30 s). This was repeated for each of the selected target whiskers (maximum 4). Experimental sessions varied in terms of electrode placement, stimulated whiskers, stimulation frequencies and number of trials per unique stimulus. 7 experimental sessions used single shanks with 16 electrodes, whilst 8 experimental sessions used 8 neighbouring shanks (16 electrodes per shank) covering 2-3 neighbouring barrels. Spike sorting classified units by E/I type and cortical layers L2/3, L4, L5A, L5B (see Methods \ref{methods:spike_sorting}).

@@

The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering relatively controlled study of the neural representation of atomic entities. Excluding non-spiking responses over 25 and 100ms following cortical onset (5.2ms following stimulus onset; Supplementary Fig. 1), responses to single whisker deflections in barrel cortex are largely single spike responses (Fig. 2a). Spike count encoding therefore has a low capacity for discriminable stimulus encoding, made worse by high spike count variance (Fig. 2b). It is possible that double spike responses are examples of superpositions of single spike representations. The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig. 2c). The paper aims to better characterise this initial (largely single spike) response.

@@

@@

>>>>

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure1.pdf}
\caption{Explanatory Figures. 
(a) Left: For two neurons (x and y-axis), illustration of sample (dots) and boundary of underlying (circle) first spike stimulus-response distributions for two hypothetical stimuli (red and gold). Right: Illustration of corresponding relative first spike differences. 
(b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
(c) Illustration of boundary of state-dependent first spike response distribution for a single state value. The green line shows the mean points that would be predicted for a continuous set of state values.
(d) Illustration of superposition of two single spike pattern feature representations.
(e) Illustration of `stationary clustered time-warped precise spike timing'. Left: Responses form several clear clusters. The main green cluster is correlated with non-unitary gradient (i.e. different from $45^{\circ}$). Under the assumption of stationarity, the green line and circle are the same as in (c). The orange ellipses illustrate the boundary of an underlying response distribution under the assumption of independence and normality for the sample (solid) and another stimulus (dashed). Centre: Illustration of spike time differences for values of the state. The green circle illustrates the boundary of a state-dependent distribution of spike time differences. Right: Conceptual illustration of time warp relative differences.}
\label{fig:fig1}
\end{figure}

@@

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}. The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities. The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information. Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units. That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is strongly recommended to view these plots in the top left of Supplementary Vid. 1. Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.

@@

Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing. Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.

A novel algorithm extracted the observed clusters and found that they were correlated above chance levels. To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity. The cluster spike times of both units were stationary for X\% of correlated clusters. X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig. 1d).

For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units. X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.

Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions. This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends. It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e. independently of state [Vicente Refs]).

The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.

The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner.

The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics. In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.

% To our knowledge this is the first report demonstrating that

% Algorithms for detecting correlated clusters and estimating correlation angles were developed for this study and verified on surrogate control datasets.

@@

% The paper uncovers interesting correlative properties, which offer a novel description of neural representations and are suggestive (but not confirmatory) of an internal continuous state varying responses from trial-to-trial.

@@

% \textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?

% Firstly, for repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity), are the first spike times of pairs of units correlated? This analysis should take into account outliers and the possibility that responses come .

% In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels. Clustering was necessary for the detection of a large proportion of correlations.

% \textit{2.} For sets of correlated spike pairs, are the angles of correlation significantly different from $45^{\circ}$ (i.e. is the gradient of correlation different from 1)? This question is of interest for two reasons. Firstly such . This leads to the

% \textit{3.} Does this offer a potential improvement in encoding capacity

% \textit{4.} For pairs of units in which

% we addressed the following questions / hypotheses. (1) …. (2) …. (3)…

% Although the study analyses responses to single stimuli in isolation, it also considers the proposed and previously suggested neural codes with respect to feature-binding and continua encoding.

@@

The dataset used for this paper is reused from an earlier study, of which analyses and data collection methods have previously been reported \cite{reyes2014laminar}. In summary, extracellular recordings were made in 15 anaesthetised Wistar rats using 16 or 128 electrode silicon probes, inserted perpendicularly into the barrel cortex. Single whisker deflections were made at various frequencies (0.1-10Hz) in blocks, each containing typically 200 trials (sweeps) with an intertrial interval of 5 s (occasionally 10, 15, or 30 s). This was repeated for each of the selected target whiskers (maximum 4). Experimental sessions varied in terms of electrode placement, stimulated whiskers, stimulation frequencies and number of trials per unique stimulus. 7 experimental sessions used single shanks with 16 electrodes, whilst 8 experimental sessions used 8 neighbouring shanks (16 electrodes per shank) covering 2-4 barrel-related columns (median 4). Units were classified as excitatory or inhibitory and assigned to a cortical layers L2/3, L4, L5A and L5B/6 (see Methods \ref{methods:spike_sorting}). Spiking activity is analysed following cortical activation onset, determined as 5.2ms following stimulus onset (Supplementary Fig. 1).

@@

This section justifies the analysis of first spike patterns as a cortical representation of single whisker deflections. Firstly, this section demonstrates that the majority of neurons respond to single whisker deflections with either 0 or 1 spike (Fig. 2a). Spike count encoding therefore has a low capacity for discriminable stimulus encoding, made worse by high spike count variance (Fig. 2b). The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig. 2c). The paper aims to better characterise this initial (largely single spike) response.

@@

This section demonstrates that the responses of some neurons exhibit depressive latency adaption, which should be controlled for when testing for correlative structure caused by a non-adaptive state.

@@

It is possible that double spike responses are examples of superpositions of single spike representations, which would increase the encoding capacity further by allowing co-activity of overlapping cell assemblies.
